\chapter*{Acknowledgements}\thispagestyle{plain}
In these two pages, I would like to express my gratitude to those who supported me during my studies at Imperial College. 

My greatest thanks to my supervisors, Paul Kelly and David Ham, for the time they have spent with me. Your guidance and passion have been an incredible source of inspiration for my research. You probably do not even realize how much I have learnt from you, and for this I will be eternally grateful.

I have been extremely lucky to work next to Doru Bercea, Luigi Nardi, Florian Rathgeber, Francis Russell, Georgios Rokos, Lawrence Mitchell, Andrew McRae, Thomas Gibson, Miklós Homolya, Michael Lange. It has been a real pleasure to share the office with you. I think we have done some real science together, and all of you have contributed to my personal and professional growth. Thanks. Keep up the hard work, I am sure we will meet again!

I met Carlo Bertolli for the first time in 2008; he was a lab helper when I was a second year student in Computer Science at University of Pisa. He co-supervised my Bachelor thesis back in 2009. We kept in touch, and after a few years he suggested me to apply for a PhD position at Imperial. If I am writing these lines, it is also thanks to him and his guidance. So, thanks Carlo. 

At the very beginning of my studies at Imperial, I met J. ``Ram" Ramanujam. We have worked, shared many interesting stories (and trolls) about the Polyhedral model, and had many pizzas and coffees together. I consider you like my third, ``unofficial" supervisor. Thanks!

With Emanuele Vespa I have shared my office for only one year, but our friendship goes back to the good old days in Pisa (when we were young). I will miss you, but I am also sure we will always be in touch. I will really miss two things: 1) our (self-)trolling, especially for all those PhD students telling us that they had deadlines and were publishing stacks of papers everywhere, while we were spending most of our time chasing bugs and ``understanding what to do next"; 2) playing Teeworlds. 

I also thank my Italian friends, those in London (in particular, Francesca, Simone, Federico), those spread all over the world, and those who I talk with almost everyday. I am so glad that after all these years we still keep in touch. 

Finally, my biggest thanks goes to those who have made this experience possible: Alice and our families. I like science, I love what I do, but without your love and support this PhD would have never been possible. My gratitude for these 4 years in London will be eternal.

